---
title: "bayesian-transmission"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bayesian-transmission}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bayestransmission)
```
# Introduction
This package provides a Bayesian framework for transmission modeling on an
individual patient level. Modeling is conducted through Markov Chain Monte
Carlo (MCMC) methods.This document will explain the basic usage of the package,
specification of parameters, and the output of the model.

# Data Structure
The algorithms expect a longitudinal data set with the following columns:
* `facility`: The facility where the event occurred.
* `unit`: The unit within the facility where the event occurred.
* `time`: The time at which the event occurred.
* `patient`: The patient involved in the event.
* `type`: The type of event.

The package includes a simulated dataset, `simulated.data`. 
```{r}
pillar::glimpse(simulated.data)
```

There are 12 different types of events that can be specified in the `type` column.
These are, expected numerical codes shown in parentheses:

* Admission (0) and Discharge (3)
* Surveillance Testing Results
    + Negative Test (1)
    + Positive Test (2)
* Clinical Testing Results
    + Negative Test (4)
    + Positive Test (5)
* Generic Testing
    + Negative Test (7)
    + Positive Test (8)
* Antibiotic Use
    + single dose (9)
    + Start (10)
    + Stop (11)
* Isolation Procedures
    + Start (6)
    + Stop (7)

Not all events need to be used in every data set, but the model selected should
reflect the data that is available. Care should be taken to correctly code the
data. The `EventToCode` and `CodeToEvent` functions can be used to convert 
between.


```{r}
table(CodeToEvent(simulated.data$type))
```



# Model Specifiction

## Model Choice
Since the model is implemented in C++ for speed and efficiency, only the
specified models can be used.  The currently implemented models are:

* `"LinearAbxModel"`, A Linear model with antibiotic use as a covariate.
* `"MixedModel"`, and
* `"LogNormalModel"`

Model specification and all parameters are controlled through constructor functions of the same name, or generically through the `LogNormalModelParams()` function.

For all models there is the choice of either a 2 state (susceptible and colonized) or 3 state (susceptible, colonized, and recovered or latent) model.  Number of states is set through the `nstates` parameter, and the number of states in the model overrides what may be specified in any individual component.

## Parameters

The remainder of the parameters are grouped into the following categories:

* `Abx`, Antibiotic use,
* `AbxRate`, Antibiotic rates,
* `InUnit`, In unit infection rates,
* `OutOfUnitInfection`, Out of unit infection rates,
* `Insitu`, In situ parameters,
* Testing:
    + `SurveilenceTest`, Surveillance testing,
    + `ClinicalTest`, Clinical testing.

Unless otherwise specified the parameters are all distributed gamma  with
specified shape and rate parameters.  Each parameter can also be left as fixed
or be sampled at each iteration of the MCMC.

### Specifying parameters

Parameters for the model may be specified by the `Param()` function. 
This function takes up to four arguments:
    
    1. `init`, is the initial value of the parameter.
    2. `weight`, is the weight of the prior distribution in updates.
    3. `update`, a flag of if the parameter should be sampled in the MCMC algorithm. 
        `FALSE` indicates that the parameter should be fixed, and is by default `TRUE` when `weight` is greater than zero.
    4. `prior`, the mean of the prior distribution.  Taken with the weight will fully parameterize the distribution.
    




## `Abx` Antibiotic use

Antibiotic use is specified by the `Abx` parameter.  This parameter is a list constructed with the `AbxParams()` function with the following components:

* `onoff`, If antibiotics are being used or not.  The two following parameters are only used if `onoff` is `TRUE`.
* `delay`, the delay for the antibiotic to take effect.
* `life`, the duration where the antibiotic to be effective.

```{r}
abx <- AbxParams(onoff = TRUE, delay = 0.3, life = 1)
```


Currently, all antibiotics are assumed to be equally effective and have the same duration of effectiveness.

## `AbxRate` Antibiotic rates

The `AbxRate` parameter control the antibiotic administration rates.

```{r}
abxrate <- AbxRateParams(uncolonized = 0.05, colonized = 0.7)
```

Here since both parameters are non-zero both will be updated. A rate of zero for 
either would indicate that group would never be on antibiotics.

## `InCol` In unit infection rate

Transmission within unit is the main defining characteristic that differentiates
models.  For example the linear antibiotic model, `LinearAbxModel()`, is differentiated from the log 
normal model, `LogNormalModelParams()` by the use of a `LogNormalABXInUnitParameters()` for the 
`InUnit` argument rather than the `LogNormalInUnitAcquisition()` which does not 
take into account antibiotic use.  All in unit transmission is defined in terms
of acquisition, progression, and clearance. 

### Aqcuisition Model
In the base log normal antibiotic model, `LogNormalABXInUnitParameters()` log acquisition probability is a linear function.

$$
\log(P(\mathrm{Acq(t)})) = \beta_0 + \beta_t(t-t_0) + 
    \beta_c N_og(\mathrm{Colonized(t)})a_{ca}N_{ca}(t) + \beta_A A_i(t) + \beta_E E_i(t)
$$
Where $\beta_\star$ represents the coefficient corresponding to the amounts, $N_c(t)$ represent the total number of colonized patients at time $t$, $N_{ca}(t)$ the number of colonized on antibiotics, and $A_i(t)$ and $E_i(t)$ represents if patient $i$ is currently or ever on antibiotics.

The linear antibiotic (`LinearAbxAcquisitionParams`) takes a more complicated form for the acquisition model.
$$
P(\mathrm{Acq(t)}) =
  \left[e^{\beta_\mathrm{time}(t-t_0)}\right]\\
  \left\{e^{\beta_0}
         \left[
            \left(\frac{\beta_\mathrm{freq}}{P(t)}+(1 - e^{\beta_\mathrm{freq}})\right)
            e^{\beta_\mathrm{mass}}\left(
                (N_c(t) - N_{ca}(t)) + e^{\beta_\mathrm{col\_abx}}N_{ca}(t)
            \right)
          + 1 - e^{\beta_\mathrm{mass}}
        \right]
  \right\}\\
  \left[
    N_S(t) - N_E(t) + e^{\beta_\mathrm{suss\_ever}}\left(\left(E_i(t)-A_i(t)\right) +A_i(t)e^{\beta_\mathrm{suss\_abx}}\right)
  \right]
$$

```{r}
progression <- LinearAbxAcquisitionParams(
    base = Param(0, 1, TRUE),
    time = 1,
    mass = 
    
)
```

### Progression Model

In the 3 state model there is a latent state and the progression model controls 
how patient transition out of the latent state.  The progression model is simple
for the log normal antibiotic case.

$$
\log(P(\mathrm{progression})) = \delta_0+\delta_AA_i(t) + \delta_EE_i(t)
$$
The linear antibiotic model is:
$$
P(\mathrm{progression}) = e^{\delta_0}\left[1-E_i(t)+e^{\delta_2}\left(E_i(t)-A_i(t)+e^{\delta_1}A_i(t)\right)\right]
$$

Where here we use $\delta$ for the coefficients, but the notation is the same.

### Clearance Model

The clearance model is the same as the progression model in both
the log normal and the linear cases, the coefficients however are independent.


## Out of Unit Importation

The out of unit parameters control the rate admission come in colonized, or in 
each stage, un a multistage model.


## In Situ

**TODO**

I'm not sure what these parameters do. It's a set of gamma distributed 
parameters one for each state.  The updates and probabilities are not time 
dependent.

When updating the rates for each state are sampled from a $gamma(N_i, 1)$ distribution.  Then all three are normalized to sum to 1.

## Testing
There are two types of testing, surveillance, which is conducted routinely at 
regular intervals such as on admission then every 3 days after, and clinical, 
where the testing is precipitated by staff, and thus the timing is informative.


### Surveillance Testing

The timing of surveillance testing is assumed to not be informative.  Therefore,
surveillance testing is only parameterized in terms of probability of a positive test given the underlying status.
Surveillance test parameters are updated with a sample from a $Beta(N_{s,1}, N_{s,0})$ distribution where $N_{s,1}$ and $N_{s,0}$ are the number of positive and negative tests respectively for state $s$.

### Clinical Testing

Since clinical testing time is informative, clinical testing is assumed to be at 
random within infection stage.  The rate of testing within each stage is sampled
from a gamma distribution. Sensitivity/Specificity are handled the same as 
surveillance testing and the likelihood is multiplicative between rate and 
effectiveness.















